---
title: "Assignment 1"
subtitle: "DABN19: Data Visualization"
author: "Dirk Baars"
date: "`r Sys.Date()`"
papersize: a4
geometry: margin=4cm
colorlinks: true
output:
  pdf_document:
    number_sections: true
---

```{r setup, include = FALSE}
# Setup options for R Markdown
knitr::opts_chunk$set(
  echo       = TRUE,    # Do not print code
  warning    = FALSE,    # Suppress warnings
  message    = FALSE,    # Suppress messages
  fig.align  = "center", # Center figures
  fig.width  = 2.7,      # Good standard figure width for single-panel figures
  fig.height = 2.4       # Good standard figure height for single-panel figures
)

library(tidyverse)

# Set a theme for ggplot2
theme_set(theme_grey(base_size = 10))

# Set options
options(
  digits = 3, # limit the number of significant digits
  width  = 63 # limit the width of code output
)
```

# Task 1

First, the data is filtered, whereby only Tatooine homeworld is 
chosen, via the filter() command. Then, the missing cases are 
deleted via the drop_na() command, and finally the data is grouped 
by eye color via grouped_by(), so finally the average mean is calculated 
via the summarize function per group.

```{r bar chart1, fig.cap= 'Average weight of Tatooine character by eye color'}
starwars_tatooine_summary <- 
  starwars %>%
  filter(homeworld == "Tatooine") %>%
  drop_na(mass) %>%
  group_by(eye_color) %>%
  summarize(avg_mass = mean(mass))

ggplot(starwars_tatooine_summary, aes(eye_color, avg_mass)) +
  geom_col() +
  xlab("Eye Color") +
  ylab("Average Weight")

```







# Task 2

2.1 The data in table4a is not tidy, because each
variable does not have its own column (year and tuberculosis cases)
and each observation does not have its own row.
2.2 In table 4a there are the following variables: Country, year and cases of tuberculosis. 

To make the data tidy, it is pivoted in a longer way by using pivot_longer(),
whereby the column names are merged into one variable 'year' 
and the values all got their own row under the column 
cases.tuberculosis and are separated by years.

```{r table1, echo=TRUE}
table4a_tidy <- table4a %>%
  pivot_longer(
    cols = c("1999", "2000"),
    names_to = "year", 
    values_to = "cases.tuberculosis"
  )

knitr::kable(table4a_tidy, digits = 5, format.args = list(scientific=
FALSE,big.mark=","),caption = 
'Tuberculosis cases in three countries across two years')
```

# Task 3

Finally, table 4b is also made tidy in the same matter
as table 4a. Then the two are joined via left_join().
Since the column names are similar, no 'by =' is used.
After that a new column is created via mutate(), 
calculating the rate of tuberculosis.

```{r table2, echo=TRUE}
table4b_tidy <- table4b %>%
  pivot_longer(
    cols = c("1999", "2000"),
    names_to = "year", 
    values_to = "population"
  )

table4 <- left_join(table4a_tidy, table4b_tidy) %>%
  mutate(rate = cases.tuberculosis/population )

knitr::kable(table4, digits = 5, format.args = list(scientific=FALSE, 
big.mark=","),caption = 'Tuberculosis cases in 
three countries across two years, including 
populations numbers and tuberculosis rate.')
```


