---
title: "Assignment 3"
subtitle: "STAE04/DABN19: Data Visualization"
author: "Dirk Baars"
date: "`r Sys.Date()`"
papersize: a4
geometry: margin=4cm
colorlinks: true
output:
  pdf_document:
    number_sections: true
---

# Task 1

The data set contains panel data of 595 observations, that was collected between 1976 and 1982 in the United States. The total numbers of observations are: 4165. The Wages data set contains 12 variables, whereby one variable is added, by taking the inverse log of lwages. 

The Description of the variables are as follows: 

| Name    | Transformation| Measurement | Description                      |
| :---    | :---          | :---        | :---                             |
| exp     | ---           | years       | years of full-time experience    |
| wks     | ---           | weeks       | weeks worked                     |
| bluecol | ---           | yes, no     | blue collar?                     |
| ind     | ---           | yes, no     | works in a manufacturing industry ?  |
| south   | ---           | yes, no     | resides in the south ?           |
| smsa    | ---      | yes, no | resides in a standard metropolitan statistical are ?  |
| married | ---           | yes, no     | married ?                        |
| sex     | ---           | male, female| sex of the person                |
| union   | ---           | yes, no     | individual’s wage set by a union contract ? |
| ed      | ---           | years       | years of education               |
| POC     | ---           | yes, no     | is the individual person of color ? |
| lwage   | log transformation| logarithm  | logarithm of wage             |
| wage    | ---           | ---         | wage in month                    |
```{r, include = FALSE}

library(tidyverse)
library(Ecdat)
library(ggmosaic)
Wages1 <- Wages %>%
  mutate(wage = exp(lwage)) 

Wages1 %>%
  ggplot(aes(exp, wage, alpha = 0.2)) +
  geom_point() +
  geom_jitter(position = position_jitter(width = 0.2, height = 0))

```



The graph below shows a relation between the years of experience, and the wage per month. A geom_point() is used to visualize the graph.To prevent overlapping, the position is jittered, and opacity (alpha = 0.5) is adjusted. Also, a logarithmic scale is adjusted to give a better overview of the data. Also, a line is added by using geom_smooth(). The graph shows that until approximately 30 years of working experience, the wage seems to increase, and after that the wages slowly decreases. 
```{r, fig.cap = "The relation between wages and years of experience. Points are jittered", fig.width = 5, fig.height = 3}

  Wages1 %>%
  ggplot(aes(exp, wage, alpha = 0.5)) +
  geom_point(position = position_jitter()) +
  labs(
    x = "Years of Work Experience", 
    y = "Wage per Month in USD"
  ) + 
  geom_smooth() +
  scale_y_log10() 
 

```

# Task 2

The graph below shows a relation between the logarithmic wage, and the weeks worked. The weeks worked is filtered, so only >=40 weeks is taken into account. A geom_point() is used to visualize the graph.To prevent overlapping, the position is jittered, and opacity (alpha = 0.5) is adjusted. Also, a the logarithmic wage is used to give a better overview of the data. Also, a line is added by using geom_smooth().

In the figure it shows a pattern, whereby wage remains fairly stable in comparison to the weeks worked. 

```{r, fig.cap = "The amount of weeks worked has a positive relation to wage. Points are jittered", fig.width = 5, fig.height = 3}
Wages1 %>%
  filter(wks >= 40) %>%
  ggplot(aes(wks, lwage, alpha = 0.5)) +
  geom_point(position = position_jitter(width = 1, height = 1)) +
  geom_smooth() + 
  labs(
    x = "Weeks Worked",
    y = "Log of montly Wage in USD"
  ) 


```

# Task 3

```{r, include = FALSE}
Wages2 <-
  Wages1 %>%
  mutate(wage_levels = cut(
    wage,
    c(0, 750, 1500, max(wage)),
    labels = c("Low", "Mid", "High")
  ))
```


When plotting sex against wage levels in a mosaic table, it is visible that most people are earning a low level of wage per month. then regarding proportion, with every level of wage, females are declining in participation of level.  
```{r, fig.cap = "Wage levels per sex", fig.width = 5, fig.height = 3}

Wages2 %>%
  ggplot() +
geom_mosaic(aes(x=product(wage_levels), fill = sex)) +
  labs(
    x = "Wage levels", 
    y = "Sex"
  )

```

```{r, include = FALSE}
Wages2 %>%
  ggplot(aes(wage_levels, fill = sex)) +
  geom_bar(position="dodge")
```





